RegressionPvNP

### 2017-07-05 RegressionPvNP.Rmd
...

---
title: "RegressionPvNP"
author: "Sarah Killcoyne"
date: "23 May 2017"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
library(biomaRt)
library(ggplot2)
library(ggfortify)
library(GGally)
library(plyr)
library(pander)
library(reshape2)
library(gridExtra)
library(GenomicRanges)
library(glmnet)

load("Test_patients.Rdata", verbose=T)
validation.patient.data = patient.data
length(validation.patient.data)

dataset = 'Training'
load(paste(dataset, '_patients.Rdata', sep=''), verbose=T)

@@

@@

## Patient info file
patient.file = grep('All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

patient.info = read.patient.info(patient.file, set=dataset)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

validation.patient.info = read.patient.info(patient.file, set='Test')
validation.patient.info = arrange(validation.patient.info, Status, Patient, Endoscopy.Year, Pathology)
sum.validation.info = summarise.patient.info(validation.patient.info)

@@

@@

pander(sum.validation.info[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')], justify='left', caption='Validation patient set')

@@

As suggested by Moritz it is possible to merge all of the patient samples and use GLMs to model the differences, and to select regions that may be most reflective of the differences. In this report that's what I've done, and it worked very well.  However, I'm a bit suspicious that it's worked too well.  I'm looking for biases in the data, or mistakes I might have made in setting up the model.

@@

```{r tile, echo=F, message=F, warning=F, include=F}
#tile.w=1e07
tile.w=5e06
tile.files = grep(as.character(tile.w), analysis.files, value=T, fixed=T)
load(tile.files[grep('Test|validation', tile.files)], verbose=T)
mergedDfValidation = mergedDf
dim(mergedDfValidation)

@@

cache.dir = paste(data, 'Analysis',sub('\\+', '', tile.w), sep='/')
if (!dir.exists(cache.dir)) dir.create(cache.dir)

#validatioMergedDf = tile.sample.segments(paste(as.character(tile.w),'validation_tiledpts.Rdata', sep=''), validation.patient.data)

@@

validation.df = t(mergedDfValidation[, grep('^D', colnames(mergedDfValidation))])
dim(validation.df)
```

@@

```{r cvglmnetPNP, echo=F, warning=F, message=F, eval=F}
## cv.glmnet Ridge vs Lasso
#With all `r ncol(dysplasia.df)` regions of the genome, first check if/which regression may be appropriate. It appears that pure lasso may provide the best balance for number of features with non-zero coefficients.

plot.multi.cv.glmnet<-function(x,y,alpha=c(0,1), family="binomial", title='cv.glmnet') {
  plots <- list(  )
  cv.models = data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('class','lambda','cvm'))))
  for (a in alpha) {
    cv = cv.glmnet(x, y, alpha=a, family=family)
    cv.models = rbind(cv.models, cbind('class'=paste('cv',a,sep=''), 'lambda'=log(cv$lambda), 'cvm'=cv$cvm))
    plots[[as.character(a)]] = autoplot(cv, main=paste('alpha',a), ylab=cv$name)
  }
  cv.models[c('lambda','cvm')] = lapply(cv.models[c('lambda','cvm')], function(x) as.numeric(as.character(x)))

  gg = ggplot(cv.models, aes(x=lambda, y=cvm, color=class)) + geom_point() + geom_line() + 
    labs(x='log(Lambda)', y='Binomial Deviance', title=title)

  plots[['all']] = gg

  return(plots)
}

gp = plot.multi.cv.glmnet(x=dysplasia.df, y=labels, alpha=c(0,0.5,0.8,1), title='cv.glmnet P vs NP')

cv1 = cv.glmnet(dysplasia.df, labels, alpha=1, family='binomial')
gg = autoplot(cv1$glmnet.fit, xvar='lambda', main='alpha=1 glmnet.fit from cv.glmnet') + theme(legend.position='none')

gp[['coef.1']] = gg

do.call(grid.arrange, c(gp, ncol=2, nrow=3))

@@

  for (i in 2:n) {
    s = sample(rep(seq(5), length = length(unique(pts$Patient))))
    patients = merge(patients, cbind('Patient'=unique(patients$Patient), 'group'=s), by="Patient")  
    foldcol = grep('group',colnames(patients))
    colnames(patients)[foldcol] = paste('fold',i,sep='.')
    patients[[paste('fold',i,sep='.')]] = check.sets(patients, paste('fold',i,sep='.'))
  }
  return(patients)
}

@@

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2, select='deviance') {
  if (nfolds > 5) minR = 0.1
  x = as.matrix(x)
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()
  tpts = create.patient.sets(pts, nfolds, splits, minR)
  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  cv.binomial.deviance = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(tpts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV <- glmnet(training, y[-test.rows], lambda=lambda, alpha=a, family='binomial') 
      # autoplot(fit) + theme(legend.position="none")

@@

```{r xvalpt, message=F, warning=F, echo=T, fig.height=16, fig.width=16}
nl = 1000

@@

file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
  message(paste("loading", file))
  load(file, verbose=T)
} else {
  for (a in alpha.values) {
    fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=nl, family='binomial') # all patients
    #autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

    l = fit0$lambda
    if (a == 0) {
      l = sort(c(fit0$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }
    cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=l, pts=pts, a=a, nfolds=folds, splits=splits, fit=fit0, select='deviance')

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, lambda.1se == T)
  }
  save(plots, coefs, performance.at.1se, file=file)
}

@@

```{r xvalpt2, message=F, warning=F, echo=T}
select.alpha = as.character(0.9)

performance.at.1se = do.call(rbind, performance.at.1se)
performance.at.1se$alpha = rownames(performance.at.1se)

ggplot(performance.at.1se, aes(alpha, mean)) + geom_text( aes(label=round(mean, 3)), hjust=-0.5) +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme), color='grey') + geom_point() + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('All samples,', folds, 'folds,', splits, 'patient splits'))

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('>=75%'=length(which(cf >= 0.75)),  '>=50%'=length(which(cf >= 0.5)),  '>=25%'=length(which(cf >= 0.25)), 'Total'=length(cf))
}))
rownames(stability) = names(coef.stable)

pander(stability, justify='left', caption='Number of features stable in n% or more folds')
```

@@

### Validation Data

TBD

```{r, message=F, warning=F, echo=F, eval=F}

fitV <- glmnet(dysplasia.df, labels, alpha=select.alpha, nlambda=nl, family='binomial') # all patients
#autoplot(fitV, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none') 
cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=fitV$lambda, pts=pts, a=select.alpha, nfolds=folds, splits=splits, fit=fitV)
lambda.opt = ifelse( length(cv.patient$lambda.1se) > 0, cv.patient$lambda.1se, cv.patient$lambda.min )

statusV = unlist(sapply(validation.patient.data, function(df) {
  df$info = arrange(df$info, Endoscopy.Year, Pathology)
  label = as.integer(df$info$Status == 'P') #as.integer(df$info$Pathology %in% c('HGD', 'IMC'))
  names(label) = df$info$Samplename
  return(label)
}))
names(statusV) = sub('.*\\.', '',  names(statusV))

# Progressors - by sample
length( which(pred[ names(which(statusV == 1)), ] > 0.5) )/length(which(statusV == 1))

# Non Progressors - by sample
length( which(pred[ names(which(statusV == 0)), ] < 0.5) )/length(which(statusV == 0))

@@

`r length(unique(unlist(lapply(coef.stable, function(cf)  names(which(cf > 0.8)) ))))` features are shared at all values of alpha and were non-zero in at least 80% of the folds.

```{r echo=F, warning=F, message=F, eval=F}
### Features in samples
## TODO Look at the coefficients, are the high pos ones "gains" and the low neg ones "losses" etc
plot(coefs[[select.alpha]][,1])

@@

Using the features from the alpha=0.9 models

@@

features.in.prog = dysplasia.df[names(which(labels == 1)),topF]
features.nonprog = dysplasia.df[names(which(labels == 0)),topF]

@@

# Gain or loss not correlated in prog or non

gl = cbind.data.frame(
  'P.gain' = apply(features.in.prog, 2, function(x) sum( as.integer(x >= 1.1) ) ), 
  'P.loss' = apply(features.in.prog, 2, function(x) sum( as.integer(x <= 0.9) ) ),
  'NP.gain' = apply(features.nonprog, 2, function(x) sum( as.integer(x >= 1.1) ) ),
  'NP.loss' = apply(features.nonprog, 2, function(x) sum( as.integer(x <= 0.9) ) ) )

allgl = cbind.data.frame(
  'ALL.P.gain' = apply(dysplasia.df[names(which(labels == 1)),], 2, function(x) sum( as.integer(x >= 1.1) ) ), 
  'ALL.P.loss' = apply(dysplasia.df[names(which(labels == 1)),], 2, function(x) sum( as.integer(x <= 0.9) ) ),
  'ALL.NP.gain' = apply(dysplasia.df[names(which(labels == 0)),], 2, function(x) sum( as.integer(x >= 1.1) ) ), 
  'ALL.NP.loss' = apply(dysplasia.df[names(which(labels == 0)),], 2, function(x) sum( as.integer(x <= 0.9) ) ) )

@@

topFSegments = do.call(rbind,  strsplit( topF, ':|-' ))
colnames(topFSegments) = c('chr','start','end')
topFSegments = as.data.frame(apply(topFSegments, 2, as.numeric))
topFSegments$chr = factor(topFSegments$chr, levels=chromInfo$chr)
topFSegments = arrange(topFSegments, chr, start)

@@

grid.arrange(ggplot(topFRegions, aes(reorder(chr, -adj.info.content), totalbp/length, fill=ai.cut)) + ylim(0, 0.5) +
  labs(y='Feature bp/total chr length', title='Ratio of covered base pairs', x='chr') +
  geom_bar(stat='identity') +  geom_text(aes(label=round(totalbp/length, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom'), 
  bottom='Ratio of base pair coverage per chromosome in the top non-zero features selected from the model.\nChromosomes are sorted and binned by their gene (e.g. information) content rather than by overall length.')

@@

```{r, warning=F, message=F}
mart = biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org")
biomart = useDataset("hsapiens_gene_ensembl", mart)

@@

bm = getBM(mart=biomart, attributes=attr, 
           filters=list('chromosomal_region'=sub('-', ':',topF), "biotype"=c("IG_C_gene","IG_D_gene","IG_J_gene","IG_V_gene","protein_coding","TR_C_gene","TR_D_gene","TR_J_gene","TR_V_gene")))
#head(bm)

@@

# Not related to the length of the segment
pander(with(topFSegments, cor.test(genes, (end-start))), caption="Number of genes per segment is not related to the segment length.")

topFRegions = merge(topFRegions, (bm %>% group_by(chromosome_name) %>% summarise('gene.cnt'=length(unique(hgnc_symbol)), 'exon.cnt'=sum(exon_count))), by.x='chr', by.y='chromosome_name', all.x=T)

#with(topFRegions, cor.test(protein_coding, exon.cnt))
#with(topFRegions, cor.test(gene.cnt, exon.cnt)) # expected

# Exons correlate with the adjusted information content measure, expected
#with(topFRegions, cor.test(exon.cnt, adj.info.content))

grid.arrange(ggplot(topFRegions, aes(reorder(chr, -adj.info.content), gene.cnt/protein_coding, fill=ai.cut)) + geom_bar(stat='identity') +
  labs(y='Gene Ratio', x='chr', title='Feature genes vs all protein coding genes') +
  geom_text(aes(label=round(gene.cnt/protein_coding, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom'), 
  bottom='Ratio of genes found within the boundaries of the top features vs all genes on the chromosome.\nChromosomes are ordered and binned by adjusted gene (e.g. information) content rather than length.')

>>>>

---
title: "RegressionPvNP"
author: "Sarah Killcoyne"
date: "22 June 2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 5
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
library(biomaRt)
library(ggplot2)
library(ggfortify)
library(GGally)
library(plyr)
library(pander)
library(reshape2)
library(gridExtra)
library(GenomicRanges)
library(glmnet)
library(CoxHD)
library(mice)

@@

includeDemo = FALSE

@@

load(grep('Test_patients.Rdata', analysis.files, value=T), verbose=T)
validation.patient.data = patient.data
length(validation.patient.data)

dataset = 'Training'
load(grep('Training_patients.Rdata', analysis.files, value=T), verbose=T)

## Patient info file
patient.file = grep('All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))
demo.file = grep('Demographics_full.txt', data.files, value=T)

patient.info = read.patient.info(patient.file, demo.file, set=dataset)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

validation.patient.info = read.patient.info(patient.file, demo.file, set='Test')
validation.patient.info = arrange(validation.patient.info, Status, Patient, Endoscopy.Year, Pathology)
sum.validation.info = summarise.patient.info(validation.patient.info)

@@

#ct = sapply(patient.data, function(pd) nrow(pd$seg.vals) )
#hist(ct)

#lens = sapply(patient.data, function(pd) mean(pd$seg.vals$end.pos - pd$seg.vals$start.pos) )
#hist(lens)

#vrs = unlist(lapply(patient.data, function(pd) apply(as.matrix(pd$seg.vals[,grep('^D', colnames(pd$seg.vals))]), 2, var) ))
#hist( vrs )
#pd$seg.vals[,grep('^D', colnames(pd$seg.vals))] - apply(pd$seg.vals[,grep('^D', colnames(pd$seg.vals))], 2, mean)

@@

pander(sum.validation.info[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')], justify='left', caption='Validation patient set')
```

@@

# Note about the data

This report uses all of the genomic data as well as encoded demographic information for age at diagnosis, BMI, p53 staining status, smoking, gender, length and circumference of the Barrett's segment.  None of these appear to improve the model.  This may not be surprising as most of these were controlled for in generating the cohort.

@@

```{r tile, echo=F, message=F, warning=F, include=F}
tile.w=5e06
tile.files = grep(as.character(tile.w), analysis.files, value=T, fixed=T)
load(tile.files[grep('Test|validation', tile.files)], verbose=T)
mergedDfValidation = mergedDf
dim(mergedDfValidation)

@@

cache.dir = paste(data, 'Analysis',sub('\\+', '', tile.w), sep='/')
if (!dir.exists(cache.dir)) dir.create(cache.dir)
```

@@

validation.df = t(mergedDfValidation[, grep('^D', colnames(mergedDfValidation))])
dim(validation.df)

```{r addDemoCols, echo=T, message=F, warning=F}

  ## Some sort of complexity score
  #plot(unlist(complexity.measures), col=sum.patient.data$Status)
  # t.test(
  # complexity.measures[subset(sum.patient.data, Status == 'P')$Patient],
  # complexity.measures[subset(sum.patient.data, Status == 'NP')$Patient]
  # )
  score.cx <-function(pd) {
  #  print(pd$info)
    len.var = (var(pd$seg.vals$end.pos - pd$seg.vals$start.pos))/1e12
    extreme.values = max( apply(as.matrix(pd$seg.vals[,grep('^D', colnames(pd$seg.vals))]), 2, function(col) {
      length(which(col >= (median(col)+2*sd(col)))) + length(which(col <= (median(col)-2*sd(col)))) 
    }) )

    (len.var + extreme.values)
  }

  complexity.measures = sapply(patient.data, score.cx)

  dysplasia.df = cbind(dysplasia.df, 'cx.measure'=NA)
  for (sample in rownames(dysplasia.df)) {
    dysplasia.df[sample, 'cx.measure'] = complexity.measures[subset(patient.info, Samplename == sample)$Patient]
  }
  # norm to 1 to keep within the magnitude of the genomic data
  cx = dysplasia.df[,'cx.measure'] 
  dysplasia.df[,'cx.measure'] = (log(cx)/log(max(cx)))
  #dysplasia.df[,'cx.measure'] = ((cx - min(cx))/(max(cx)-min(cx)))

  complexity.measures = sapply(validation.patient.data, score.cx)

  validation.df = cbind(validation.df, 'cx.measure'=NA)
  for (sample in rownames(validation.df)) {
    validation.df[sample, 'cx.measure'] = complexity.measures[subset(validation.patient.info, Samplename == sample)$Patient]
  }
  # norm to 1 to keep within the magnitude of the genomic data
  cx = validation.df[,'cx.measure'] 
  validation.df[,'cx.measure'] = (log(cx)/log(max(cx)))

if (includeDemo) {

  ## Add in demographics

  # sort by samples in the order of the matrix
  patient.info = patient.info[match(rownames(dysplasia.df), patient.info$Samplename),]

  patient.info[which(patient.info$BMI > 100), 'BMI'] = NA # I think this one is wrong

  demo.data = patient.info[,c('Sex','Circumference','Maximal', 'Age.at.diagnosis', 'BMI', 'p53.Status', 'Smoking')]
  demo.data$Sex = as.integer(demo.data$Sex)-1
  demo.data$Smoking = as.integer(demo.data$Smoking)-1
  demo.data$p53.Status = as.integer(demo.data$p53.Status)-1
  demo.cols = c('Sex','C','M','Age', 'BMI', 'p53.Status', 'Smoking')

  encode.age <- function(x) {
    if (x > 70) { 
      (x - 70)/6 
    } else if (x < 50) {
      (x -50)/6
    } else {
      0
    }
  }
  #demo.data$Age.at.diagnosis = scale(demo.data$Age.at.diagnosis)

  # Encode based on presumed distributions or unit normalize
  demo.data = demo.data %>% rowwise() %>% mutate( 'age.encoded'= encode.age(Age.at.diagnosis))
  demo.data$Age.at.diagnosis = demo.data$age.encoded
  demo.data$age.encoded = NULL
  demo.data$BMI = scale(demo.data$BMI)

  demo.data[,c('Circumference','Maximal')] = apply(demo.data[,c('Circumference','Maximal')], 2, function(x) (x-mean(x, na.rm=T))/sd(x, na.rm=T))

  ## Impute missing data
  imp = mice(as.matrix(demo.data), diagnostics = F)

  dysplasia.df = cbind(dysplasia.df, 'Sex'=NA,'C'=NA,'M'=NA,'Age'=NA, 'BMI'=NA, 'p53.Status'=NA, 'Smoking'=NA)
  dysplasia.df[,demo.cols] = as.matrix(complete(imp))

}
```

@@

@@

  for (i in 2:n) {
    s = sample(rep(seq(5), length = length(unique(pts$Patient))))
    patients = merge(patients, cbind('Patient'=unique(patients$Patient), 'group'=s), by="Patient")  
    foldcol = grep('group',colnames(patients))
    colnames(patients)[foldcol] = paste('fold',i,sep='.')
    patients[[paste('fold',i,sep='.')]] = check.sets(patients, paste('fold',i,sep='.'))
  }

  return(patients)
}

@@

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2, select='deviance', ...) {
  if (nfolds > 5) minR = 0.1
  x = as.matrix(x)
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()

  if (ncol(pts) == (nfolds+2)) {
    tpts = pts
  } else {
    tpts = create.patient.sets(pts, nfolds, splits, minR)
  }

  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  cv.binomial.deviance = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(tpts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV <- glmnet(training, y[-test.rows], lambda=lambda, alpha=a, family='binomial', ...) 
      # autoplot(fit) + theme(legend.position="none")

@@

```{r xvalpt, message=F, warning=F, echo=F, fig.height=16, fig.width=16}
nl = 1000

@@

file = paste(cache.dir, 'patient_folds.tsv', sep='/')
if (file.exists(file)) {
  sets = read.table(file, header=T, sep='\t')
} else {
  sets = create.patient.sets(pts, folds, splits, 0.2)  
  write.table(sets, quote=F, sep='\t', row.names=F, file=file)
}

file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
  message(paste("loading", file))
  load(file, verbose=T)
} else {
  for (a in alpha.values) {
    fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=nl, family='binomial', standardize=T) # all patients
    #autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

    l = fit0$lambda
    if (a == 0) {
      l = sort(c(fit0$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }
    cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=l, pts=sets, a=a, nfolds=folds, splits=splits, fit=fit0, select='deviance', standardize=T)

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, lambda.1se == T)
  }
  save(plots, coefs, performance.at.1se, dysplasia.df, file=file)
}

@@

```{r xvalpt2, message=F, warning=F, echo=T, fig.width=8, fig.height=6}

performance = do.call(rbind.data.frame, performance.at.1se)
performance$alpha = rownames(performance)

## check how often the feature(s) is selected at that lamda in each split. "stability selection"
coef.stable = lapply( coefs, function(cf) {
  sort(rowSums(cf[,-1]), decreasing=T)
})
coef.stable = lapply(coef.stable, function(cf) cf/(folds*splits))

performance = cbind(performance, do.call(rbind,  lapply(coef.stable, function(x) {
  cbind('n.Coef'=length(x), '25%'=length(which(x>=.25)), 
        '50%'=length(which(x>=.5)), '75%'=length(which(x>=.75 )), 
        '100%'=length(which(x==1)) )
})))

ggplot(performance, aes(alpha, mean)) + geom_text( aes(label=round(mean, 3)), hjust=-0.5) +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme), color='grey') + geom_point() + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value: ridge <-> lasso', y='Model classification at lambda.1se', 
       title=paste('All samples,', folds, 'folds,', splits, 'patient splits'))

@@

pander(performance[,c('n.Coef','25%','50%','75%', '100%')], justify='left', caption='Number of features stable in n% or more folds')

select.alpha = as.character(0.9)

# Genomic regions only
most.stable = grep('^\\d+', names(which(coef.stable[[select.alpha]] >= 0.75 )), value=T)

most.stable = cbind.data.frame(do.call(rbind, strsplit(most.stable, ':|-')), most.stable)
colnames(most.stable) = c('chr','start','stop', 'name')

most.stable$chr = as.integer(as.character(most.stable$chr))
most.stable = arrange(most.stable, chr, start)

@@

`r pander(most.stable[,c('chr','start','stop')], caption='Features identified as stable in 75% or more folds.')`

@@

```{r echo=F, warning=F, message=F, eval=F}
### Features in samples
## TODO Look at the coefficients, are the high pos ones "gains" and the low neg ones "losses" etc
plot(coefs[[select.alpha]][,1])
selected = coefs[[select.alpha]]

selected[which(
  selected[,1] > mean(selected[,1]) + sd(selected[,1]) | 
  selected[,1] < mean(selected[,1] - sd(selected[,1]))  
),]

## Variance in the selected features?  Not so much
coefs[[select.alpha]][,1]

v = vector(length(colnames(dysplasia.df)), mode='numeric')
names(v) = colnames(dysplasia.df)
v[rownames(selected)] = coefs[[select.alpha]][,1]

vv <- apply(dysplasia.df, 2, var)
plot(v, vv, log='y', col=(v!=0)+1)

boxplot(vv ~ (v==0), log='y', ylim=c(0.0001,1))

qqplot(vv[v==0], vv[v!=0], log='xy')

@@

Using the features from the alpha=`r select.alpha` models

@@

features.in.prog = dysplasia.df[names(which(labels == 1)), grep('^\\d+:', topF, value=T)]
features.nonprog = dysplasia.df[names(which(labels == 0)), grep('^\\d+:', topF, value=T)]

@@

# Gain or loss not correlated in prog or non
genomeCols = grep('^\\d+:', colnames(dysplasia.df), value=T)
gl = cbind.data.frame(
  'P.gain' = apply(features.in.prog, 2, function(x) sum( as.integer(x >= 1.1) ) ), 
  'P.loss' = apply(features.in.prog, 2, function(x) sum( as.integer(x <= 0.9) ) ),
  'NP.gain' = apply(features.nonprog, 2, function(x) sum( as.integer(x >= 1.1) ) ),
  'NP.loss' = apply(features.nonprog, 2, function(x) sum( as.integer(x <= 0.9) ) ) )

allgl = cbind.data.frame(
  'ALL.P.gain' = apply(dysplasia.df[names(which(labels == 1)), genomeCols], 2, function(x) sum( as.integer(x >= 1.1) ) ), 
  'ALL.P.loss' = apply(dysplasia.df[names(which(labels == 1)),genomeCols], 2, function(x) sum( as.integer(x <= 0.9) ) ),
  'ALL.NP.gain' = apply(dysplasia.df[names(which(labels == 0)),genomeCols], 2, function(x) sum( as.integer(x >= 1.1) ) ), 
  'ALL.NP.loss' = apply(dysplasia.df[names(which(labels == 0)),genomeCols], 2, function(x) sum( as.integer(x <= 0.9) ) ) )

@@

topFSegments = do.call(rbind,  strsplit( topF[grep('^\\d+:', topF)], ':|-' ))
colnames(topFSegments) = c('chr','start','end')
topFSegments = as.data.frame(apply(topFSegments, 2, as.numeric))
topFSegments$chr = factor(topFSegments$chr, levels=chromInfo$chr)
topFSegments = arrange(topFSegments, chr, start)

@@

grid.arrange(ggplot(topFRegions, aes(reorder(chr, -adj.info.content), totalbp/length, fill=ai.cut)) + ylim(0, 0.5) +
  labs(y='Feature bp/total chr length', title='Ratio of covered base pairs', x='chr') +
  geom_bar(stat='identity') +  geom_text(aes(label=round(totalbp/length, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom'), 
  bottom='Ratio of base pair coverage per chromosome in the top non-zero features\n selected from the model. Chromosomes are sorted\nand binned by their gene (e.g. information) content rather than by overall length.')

@@

```{r, warning=F, message=F}
bm = NULL
tryCatch({
  mart = biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org")
  biomart = useDataset("hsapiens_gene_ensembl", mart)

@@

  bm = getBM(mart=biomart, attributes=attr, 
           filters=list('chromosomal_region'=sub('-', ':',topF[grep('^\\d+:', topF)]), "biotype"=c("IG_C_gene","IG_D_gene","IG_J_gene","IG_V_gene","protein_coding","TR_C_gene","TR_D_gene","TR_J_gene","TR_V_gene")))
#head(bm)

@@

  # Not related to the length of the segment
  pander(with(topFSegments, cor.test(genes, (end-start))), caption="Number of genes per segment is not related to the segment length.")
}, error = function(e) {
  warning(paste("Failed to connect to Biomart service", e))
})
```

### Hazards

Using the model for alpha=`r select.alpha`.

```{r echo=F, warning=F, message=F,eval=T, fig.height=8, fig.width=8}
hazards = apply( exp(t(dysplasia.df[, rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
})
ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$label = rownames(ch)
library(ggrepel)
ggplot(ch, aes(coef, hazards, col=hazards > quantile(hazards)[['75%']])) + 
  #xlim(min(coefGR$coef)-5, max(coefGR$coef)+10) +
  geom_point(alpha=0.5) + 
  geom_text_repel(aes(x=coef+10, label=ifelse(hazards > quantile(hazards)[['75%']], label, '')),show.legend=F) +
  labs(title='Top features coef vs hazard') + theme(legend.position = 'bottom')
```

```{r echo=F, warning=F, message=F,eval=T, fig.height=12, fig.width=8}
ch = arrange(ch, -hazards)

ch$hazards = round(ch$hazards, 2)
ch$coef = round(ch$coef, 2)

melt.by.reg<-function(x) {
  r = x[['label']]

  reg = cbind.data.frame( 'value'=dysplasia.df[ , r ], labels)
  reg$labels = factor(reg$labels)
  reg$labels = revalue(reg$labels, c("1"="P", "0"="NP"))

  tt = wilcox.test(dysplasia.df[ which(labels == 0) , r ], dysplasia.df[ which(labels == 1) , r ])

  reg$pval = tt$p.value
  reg$region = r
  reg$hazard = x[['hazards']]
  reg$coef = x[['coef']]
  return(reg)
}

plot.reg<-function(x) {
  reg = melt.by.reg(x)
  pv = ifelse(tt$p.value < .001, '< .001', format.pval(unique(reg$p.value)))

#  ggplot(reg, aes(value, fill=labels) ) + geom_histogram(bins=50, color='grey') + 
#    labs(x='seg. value', title=r, subtitle=paste('haz=',x[['hazards']], ' coef=', x[['coef']], ' MW pval: ',pv,sep=''))
  ggplot(reg, aes(x=labels, y=value, group=labels, fill=labels) ) + geom_jitter(color='grey', show.legend = F) + geom_violin(show.legend = F, alpha=0.5) +  ylim(0.75, 1.2) +
    geom_hline(yintercept = median(reg$value) + sd(reg$value)) + geom_hline(yintercept = median(reg$value) - sd(reg$value)) +
    labs(x='', y='seg. value', title=r, subtitle=paste('haz=',x[['hazards']], ' coef=',x[['coef']], '\nMW pval ',pv,sep=''))
}

regions = apply(ch, 1, melt.by.reg)

do.call(grid.arrange, c(lapply(regions, function(reg) {
  se = 2
  x = rbind.data.frame('P'= cbind( 'gain' = length(which(with(subset(reg, labels == 'P'), value >= median(value) + sd(value)*se))),
                               'loss' = length(which(with(subset(reg, labels == 'P'), value <= median(value) - sd(value)*se))) ),

                  'NP' = cbind( 'gain' = length(which(with(subset(reg, labels == 'NP'), value >= median(value) + sd(value)*se))),
                                'loss' = length(which(with(subset(reg, labels == 'NP'), value <= median(value) - sd(value)*se))) ) )
  # x = rbind.data.frame('P'= cbind( 'high' = length(which(with(subset(reg, labels == 'P'), value >= 1.1))),
  #                                  'low' = length(which(with(subset(reg, labels == 'P'), value <= 0.9))) ),
  # 
  #                 'NP'= cbind( 'high' = length(which(with(subset(reg, labels == 'NP'), value >= 1.1))),
  #                                  'low' = length(which(with(subset(reg, labels == 'NP'), value <= 0.9))) ))

  cols = c('indianred3','slateblue2')
  if ( as.numeric(reg$coef[1]) < 0) cols = c('plum', 'skyblue3')
  #ggplot(melt(as.matrix(x)), aes(Var1, value, fill=Var2) ) + geom_bar(stat='identity', position='dodge') + labs(x='', y='', title=unique(reg$region)[1]) + scale_fill_manual(values=cols)

  ggplot(melt(as.matrix(rowSums(x))), aes(Var1, value, fill=Var1)) + geom_bar(stat='identity', show.legend = F)+ labs(x='', y='', title=unique(reg$region)[1]) + scale_fill_manual(values=cols)

})))
